---
title: "Change in arrival and departure timings of migratory birds with wintering latitude"
date: "2022/03/31"
author: "Karthik Thrikkadeeri"
editor_options: 
  chunk_output_type: console
bibliography: migration-timings-latitude.bib
link-citations: yes
output: 
  html_document:
    toc: true
    toc_depth: 3  
    number_sections: false  
    theme: united  
    highlight: tango  # specifies the syntax highlighting style
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(patchwork)
# library(boot)
# library(MASS)


source("scripts/functions.R")

theme_set(theme_bw())

# migration_palette <- c("#1B9E77", "#E89005", "#EF4050", "#9678B6")

knitr::opts_chunk$set(echo = FALSE, warnings = FALSE, 
                      messages = FALSE, cache.lazy = FALSE)

```

```{r data_filt, cache=TRUE}

# rawpath <- "data/ebd_IN_relFeb-2022.txt"
# senspath <- "data/ebd_sensitive_relDec-2021_IN.txt"
# groupaccspath <- "data/ebd_users_GA_relDec-2021.csv"
# 
# # processing the data for current use (34 mins)
# data_process(rawpath, senspath) 
# 
# # cleaning and filtering for data quality (7 mins)
# dataqual_filt(data, groupaccspath, maxvel = 20, minsut = 2)
# 
# # weekly reporting frequencies (<1 min)
# sp_repfreq(data)
# 
# # selecting species for this analysis
# data0 <-  sp_select(data) %>%
#   left_join(dataRF)
# 
# save(data0, file = "data/data0.RData")

load("data/data0.RData")

```

# Introduction

***This notebook is the exploratory analysis notebook for the study. Here, I explore, mainly via visualisations, interesting patterns in arrival and departure times of some migratory bird species changing with their wintering latitudes.***

## Questions

# Exploration

Explorations from before I joined BCI:

## Plain visualisation of migration timings with latitudes 

```{r 01_prelimvis, cache=TRUE, message=FALSE}

data1 <- data0 %>% 
  group_by(COMMON.NAME, LATITUDE, WEEK.MY) %>% 
  summarise(N = n(), 
            WEEK.Y = WEEK.Y) 

# getting kernel density estimates (avoid overplotting)
# creating a new data frame of 2d density grid
# fitting a model
# applying the model to the original data to estimate density at that point
dens <- MASS::kde2d(data1$WEEK.MY, data1$LATITUDE)
grid <- data.frame(with(dens, expand.grid(x,y)), as.vector(dens$z))
names(grid) <- c("xgr", "ygr", "zgr")
mod <- MASS::loess(zgr ~ xgr*ygr, data = grid, span = 0.5)

data1 <- data1 %>% 
  ungroup() %>% 
  mutate(POINTDENS = predict(mod, newdata = data.frame(xgr = data1$WEEK.MY, 
                                                       ygr = data1$LATITUDE)))

rm(list = c("dens","grid","mod"))

plot1 <- ggplot(data1, aes(WEEK.MY, LATITUDE, colour = POINTDENS)) + 
  scale_x_continuous(breaks = (seq(0,52,4))) +
  scale_y_continuous(breaks = (seq(0,40,5))) +
  coord_trans(y = "log10") +
  scale_colour_viridis_c() +
  labs(title = "Reports of migrant species varying with weeks and latitude",
       subtitle = "X-axis represents number of weeks starting from June to May") +
  geom_jitter(size = 0.8) +
  geom_vline(xintercept = 31) +
  annotate("text", x = 29, y = 10, label = "Dec", angle = 90) +
  annotate("text", x = 32, y = 10, label = "Jan", angle = 90) +
  facet_wrap(~COMMON.NAME) 

plot1

```

## Investigating arrival and departure timings

```{r 02_prelimvis, cache=TRUE, message=FALSE}

data2 <- data0 %>% 
  group_by(COMMON.NAME, LATITUDE1) %>% 
  summarise(N = n(),
            ARR = unname(quantile(WEEK.MY, probs = 0.025, type = 1)), 
            DEP = unname(quantile(WEEK.MY, probs = 0.975, type = 1))) %>% 
  pivot_longer(cols = ARR:DEP,
               names_to = "DATE.TYPE", values_to = "WEEK") %>% 
  mutate(DATE.TYPE = factor(DATE.TYPE, levels = c("DEP","ARR")))


plot2 <- ggplot(data2) + 
  theme(strip.placement = "outside") +
  geom_point(aes(x = WEEK, y = LATITUDE1, colour = DATE.TYPE)) +
  scale_colour_manual(values = c("#b22222","#006994")) +
  geom_vline(aes(xintercept = 31)) +
  geom_text(aes(x = 31, y = 10, label = "Dec"), nudge_x = -2, angle = 90) +
  geom_text(aes(x = 31, y = 10, label = "Jan"), nudge_x = 1, angle = 90) +
  facet_wrap(~COMMON.NAME) +
  scale_y_continuous(breaks = (seq(0, 40, 5))) +
  ggtitle("Variation of first arrival and last departure dates with latitude") 


```

For BRWa, the departures seem to be more synchronised than the arrivals, while it is the other way around for BBFl. But generally for both species, there seems to be a temporal lag of around a month between arrivals at the northernmost wintering grounds and at the southernmost ones.


Ashwin's thoughts:
- Distinguish between staying and passage using freq
- I find it really interesting that Greenish Warblers that span multiple latitudes in the peninsula show synchronous arrival and departure. But in the extreme north, there may be passage movement. You can check this using the previous method. I also suggest that you exclude data from the Himalayan states and from West Bengal eastwards so that we can focus exclusively on subspecies viridanus that has an extralimital breeding range (and exclude the Himalayan breeding and NE wintering subspecies trochiloides/ludlowi). Is there still some sign of passage just south of the Himalayas and in the northern plains? I suspect it will continue to show signs of slow passage through that region.