---
title: "Relationship between wintering latitude and arrival/departure timings of migratory birds"
date: "2024/08/15"
author: "Karthik Thrikkadeeri"
editor_options: 
  chunk_output_type: console
bibliography: migration-timings-latitude.bib
link-citations: yes
output: 
  html_document:
    number_sections: false  
    theme: united  
    highlight: tango  # specifies the syntax highlighting style
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(patchwork)
# library(boot)
# library(MASS)


source("scripts/functions.R")

theme_set(theme_bw())

# migration_palette <- c("#1B9E77", "#E89005", "#EF4050", "#9678B6")

knitr::opts_chunk$set(echo = FALSE, warnings = FALSE, 
                      messages = FALSE, cache.lazy = FALSE)

```

```{r data_filt, cache=TRUE}

# rawpath <- "data/ebd_IN_relFeb-2022.txt"
# senspath <- "data/ebd_sensitive_relDec-2021_IN.txt"
# groupaccspath <- "data/ebd_users_GA_relDec-2021.csv"

# load("data/ebd_IN_relJan-2022.RData")

# # processing the data for current use (34 mins) # ~ 2h on 2024
# data <- data_process(rawpath, senspath, data) 

# # cleaning and filtering for data quality (7 mins)
# data <- dataqual_filt(data, groupaccspath, maxvel = 20, minsut = 2)


# # prepare dataset with species and metrics of interest 
# prep_data_spec(data)


load("data/dataforanalysis.RData")


if(!dir.exists("02_explor_figs/")) {
  dir.create("02_explor_figs/", recursive = TRUE)
}

```

# Introduction

***This notebook is the exploratory analysis notebook for the study. Here, I explore, mainly via visualisations, interesting patterns in arrival and departure times of some migratory bird species changing with their wintering latitudes.***

## Questions

# Exploration

## Investigating arrival and departure timings

```{r, cache=FALSE, message=FALSE}

data1 <- data_full %>% 
  mutate(DAY.MY = if_else(DAY.Y > 147, DAY.Y-147, max(DAY.Y) - (147-DAY.Y))) |> 
  group_by(COMMON.NAME, LAT.BOX) %>% 
  distinct(GROUP.ID, .keep_all = TRUE) %>% 
  summarise(N = n(),
            ARR = unname(quantile(DAY.MY, probs = 0.025, type = 1)), 
            DEP = unname(quantile(DAY.MY, probs = 0.975, type = 1))) %>% 
  pivot_longer(cols = ARR:DEP,
               names_to = "DATE.TYPE", values_to = "DAY.MY") %>% 
  mutate(DATE.TYPE = factor(DATE.TYPE, levels = c("DEP","ARR"))) |> 
  # join week info
  left_join(data_full |> 
              mutate(DAY.MY = if_else(DAY.Y > 147, DAY.Y-147, max(DAY.Y) - (147-DAY.Y))) |> 
              distinct(DAY.MY, WEEK.MY),
            by = "DAY.MY")


arrdep_quan <- ggplot(data1 |> 
  filter(COMMON.NAME %in% c("Blyth's Reed Warbler", "Greenish Warbler", 
                            "Brown-breasted Flycatcher", "Brown Shrike", "Bank Swallow",
                            "Booted Eagle")),
  aes(x = DAY.MY, y = LAT.BOX, colour = DATE.TYPE)) + 
  theme(strip.placement = "outside") +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_colour_manual(values = c("#b22222","#006994")) +
  geom_vline(aes(xintercept = 217)) +
  geom_text(aes(x = 217, y = 10, label = "Dec", colour = NA), nudge_x = -14, angle = 90) +
  geom_text(aes(x = 217, y = 10, label = "Jan", colour = NA), nudge_x = 7, angle = 90) +
  # facet_grid(cols = vars(DATE.TYPE), rows = vars(COMMON.NAME)) +
  facet_wrap(~ COMMON.NAME, ncol = 1) +
  scale_y_continuous(breaks = (seq(0, 40, 5))) +
  labs(title = "Variation of first arrival and last departure dates with latitude",
       subtitle = "Using quantiles of weekly records per latitude to assign arrival and departure") +
  # coord_flip() +
  theme(legend.position = "none")

arrdep_quan

ggsave("02_explor_figs/01_arrdep_quan.png", plot = arrdep_quan,
       width = 13, height = 7, units = "in")



# palette1 <- c("#6D9C94", "#F8EF8C", "#512D38") # like a dull viridis
# palette2 <- c("#F0F2EF", "#5C5346")
# 
# 
# data3 <- data0 %>% 
#   group_by(COMMON.NAME, LAT.BOX) %>% 
#   filter(round(REP.FREQ, 2) %in% seq(0.00, 0.75, 0.01)) %>%
#   ungroup() %>% 
#   distinct(COMMON.NAME, LAT.BOX, WEEK.MY, REP.FREQ)
# 
# ((ggplot(filter(data3, COMMON.NAME == "Blyth's Reed Warbler")) +
#     geom_point(aes(x = WEEK.MY, y = LAT.BOX, colour = REP.FREQ), size = 3) +
#     scale_colour_gradientn(colours = palette2) +
#     # scale_colour_viridis_c(option = "magma", direction = -1) +
#     geom_vline(aes(xintercept = 31)) +
#     geom_text(aes(x = 31, y = 7, label = "Dec"), nudge_x = -2, angle = 90) +
#     geom_text(aes(x = 31, y = 7, label = "Jan"), nudge_x = 1, angle = 90) +
#     scale_y_continuous(limits = c(7, 37), breaks = (seq(0, 40, 5))) +
#     labs(subtitle = "Blyth's Reed Warbler")) |
# (ggplot(filter(data3, COMMON.NAME == "Brown-breasted Flycatcher")) +
#    geom_point(aes(x = WEEK.MY, y = LAT.BOX, colour = REP.FREQ), size = 3) +
#    scale_colour_gradientn(colours = palette2) +
#    # scale_colour_viridis_c(option = "magma", direction = -1) +
#    geom_vline(aes(xintercept = 31)) +
#    geom_text(aes(x = 31, y = 7, label = "Dec"), nudge_x = -2, angle = 90) +
#    geom_text(aes(x = 31, y = 7, label = "Jan"), nudge_x = 1, angle = 90) +
#    scale_y_continuous(limits = c(7, 37), breaks = (seq(0, 40, 5))) +
#    labs(subtitle = "Brown-breasted Flycatcher")) |
# (ggplot(filter(data3, COMMON.NAME == "Greenish Warbler")) +
#    geom_point(aes(x = WEEK.MY, y = LAT.BOX, colour = REP.FREQ), size = 3) +
#    scale_colour_gradientn(colours = palette2) +
#    # scale_colour_viridis_c(option = "magma", direction = -1) +
#    geom_vline(aes(xintercept = 31)) +
#    geom_text(aes(x = 31, y = 7, label = "Dec"), nudge_x = -2, angle = 90) +
#    geom_text(aes(x = 31, y = 7, label = "Jan"), nudge_x = 1, angle = 90) +
#    scale_y_continuous(limits = c(7, 37), breaks = (seq(0, 40, 5))) +
#    labs(subtitle = "Greenish Warbler"))) +
# plot_annotation(title = "Variation of first arrival and last departure dates with latitude",
#                 subtitle = "Using weekly reporting frequencies to judge passage and wintering") -> arrdep_rf
# 
# ggsave("01_prelim_figs/02_arrdep_rf_palette2.png", plot = arrdep_rf,
#        width = 15, height = 8, units = "in")
# 
# 
# 
# # should probably use proportion of max rep.freq. to colour than just rep.freq.
# 
# data4 <- data0 %>% 
#   distinct(COMMON.NAME, LAT.BOX, WEEK.MY, REP.FREQ) %>% 
#   group_by(COMMON.NAME, LAT.BOX) %>%
#   mutate(MAX.RF = max(REP.FREQ)) %>% 
#   group_by(COMMON.NAME, LAT.BOX, MAX.RF, WEEK.MY) %>% 
#   summarise(REP.FREQ = REP.FREQ,
#             PROP.RF = REP.FREQ/MAX.RF) %>% 
#   ungroup()
# 
# ((ggplot(filter(data4, COMMON.NAME == "Blyth's Reed Warbler")) +
#     geom_point(aes(x = WEEK.MY, y = LAT.BOX, colour = PROP.RF), size = 3) +
#     scale_colour_gradientn(colours = palette2) +
#     # scale_colour_viridis_c(option = "magma", direction = -1) +
#     geom_vline(aes(xintercept = 31)) +
#     geom_text(aes(x = 31, y = 7, label = "Dec"), nudge_x = -2, angle = 90) +
#     geom_text(aes(x = 31, y = 7, label = "Jan"), nudge_x = 1, angle = 90) +
#     scale_y_continuous(limits = c(7, 37), breaks = (seq(0, 40, 5))) +
#     labs(subtitle = "Blyth's Reed Warbler")) |
# (ggplot(filter(data4, COMMON.NAME == "Brown-breasted Flycatcher")) +
#    geom_point(aes(x = WEEK.MY, y = LAT.BOX, colour = PROP.RF), size = 3) +
#    scale_colour_gradientn(colours = palette2) +
#    # scale_colour_viridis_c(option = "magma", direction = -1) +
#    geom_vline(aes(xintercept = 31)) +
#    geom_text(aes(x = 31, y = 7, label = "Dec"), nudge_x = -2, angle = 90) +
#    geom_text(aes(x = 31, y = 7, label = "Jan"), nudge_x = 1, angle = 90) +
#    scale_y_continuous(limits = c(7, 37), breaks = (seq(0, 40, 5))) +
#    labs(subtitle = "Brown-breasted Flycatcher")) |
# (ggplot(filter(data4, COMMON.NAME == "Greenish Warbler")) +
#    geom_point(aes(x = WEEK.MY, y = LAT.BOX, colour = PROP.RF), size = 3) +
#    scale_colour_gradientn(colours = palette2) +
#    # scale_colour_viridis_c(option = "magma", direction = -1) +
#    geom_vline(aes(xintercept = 31)) +
#    geom_text(aes(x = 31, y = 7, label = "Dec"), nudge_x = -2, angle = 90) +
#    geom_text(aes(x = 31, y = 7, label = "Jan"), nudge_x = 1, angle = 90) +
#    scale_y_continuous(limits = c(7, 37), breaks = (seq(0, 40, 5))) +
#    labs(subtitle = "Greenish Warbler"))) +
# plot_annotation(title = "Variation of first arrival and last departure dates with latitude",
#                 subtitle = "Using proportion of max. reporting frequency per latitude to judge passage and wintering") -> arrdep_proprf
# 
# ggsave("01_prelim_figs/02_arrdep_proprf.png", plot = arrdep_proprf,
#        width = 15, height = 8, units = "in")


```

Arrival and departure timings have been obtained using quantiles (see fig). Now calculating migration metrics such as norther/southern arrival & departure, and synchronicity.

```{r, cache=FALSE, message=FALSE}

data2 <- data1 %>% 
  group_by(COMMON.NAME, DATE.TYPE) %>% 
  arrange(COMMON.NAME, DATE.TYPE, desc(LAT.BOX)) %>% 
  mutate(NORTH.DAY = first(DAY.MY),
         NORTH.WEEK = first(WEEK.MY),
         SOUTH.DAY = last(DAY.MY),
         SOUTH.WEEK = last(WEEK.MY)) %>% 
  # scaling DAY.MY between north and south 
  # mutate(DAY.SCALE = case_when(DATE.TYPE == "ARR" ~ DAY.MY - NORTH.DAY,
  #                              DATE.TYPE == "DEP" ~ DAY.MY - SOUTH.DAY))
  # mutate(DAY.SCALE = case_when(DATE.TYPE == "ARR" ~ DAY.MY - min(DAY.MY),
  #                              DATE.TYPE == "DEP" ~ DAY.MY - min(DAY.MY)))
  mutate(DAY.SCALE = scale(DAY.MY)[,1],
         LAT.SCALE = scale(LAT.BOX)[,1])

x <- data2 %>% 
  group_by(COMMON.NAME, DATE.TYPE) %>% 
  reframe(SLOPE = lm(LAT.BOX ~ 0 + DAY.MY, 
                     offset = rep(min(LAT.BOX), n()),
                     data = cur_group())$coef[1]) 
  # # truncating arrivals and departure to one direction
  # mutate(SLOPE = case_when(DATE.TYPE == "ARR" & SLOPE > 0 ~ -1,
  #                          DATE.TYPE == "DEP" & SLOPE < 0 ~ 1,
  #                          TRUE ~ SLOPE))

lm(LAT.BOX - 8 ~ 0 + DAY.MY,
   data = data2 %>% filter(COMMON.NAME == "Greenish Warbler", DATE.TYPE == "DEP"))

```

